\Name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Method for a Precision Spectrograph

### 2020-08-25 paper/paper.tex
All Comments Integrated
I even re-read the paper myself.

\documentclass[modern]{aastex63}
\usepackage[ruled,vlined,norelsize]{algorithm2e}

% hogg dealing with his shit
\addtolength{\topmargin}{-0.30in}
\addtolength{\textheight}{0.70in}
\renewcommand{\twocolumngrid}{} % hahaha!
\setlength{\parindent}{1.1\baselineskip}
\sloppy\sloppypar\raggedbottom\frenchspacing
\shorttitle{excalibur: a wavelength-calibration model}
\shortauthors{zhao, hogg, bedell, fischer}

@@

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments, which feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all possible calibration states for an instrument.  \Name\ also takes advantage of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities, better than any parametric model, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which includes a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  In a typical test case, radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.  This improvement is equivalent to removing an independent, additive error of $0.62\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity (\eprv) programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Both \expres\ and \espresso\ have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, peterburg2020, brewer2020, mascareno2020}.

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.  The move towards 0.1 $\mps$ RV precision, however, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to identify incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using a segmented polynomial in the dispersion direction, tuned to capture detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

@@

\section{Method} \label{sec:method}
The \name\ method is designed to take many calibration images, each containing a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  It operates on the two core ideas; The wavelength solution should be allowed flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \Name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to a 2D, 9\textsuperscript{th}-order polynomial, where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as, in our application, this dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ can be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$..

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present a model using principal component analysis (PCA) \citep{wiki_pca}.  A PCA is justifiable in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  We exaggerate variations in measured line position, changes in calibration space, and interpolation deviations for clarity.  In step one, dimensionality reduction and denoising (\textsection \ref{sec:denoising}), the complete set of line positions for all exposures is analyzed to return a set of $K$ basis vectors, $g_n$, which represent different ways the spectrograph calibration changes.  These basis vectors span the $K$-dimensional calibration space of the spectrograph, which includes all possible wavelength solutions.  In step two (\textsection \ref{sec:interp_time}), the amplitude of each basis vector, $a_{n,k}$, is interpolated to return the calibration state for a specific science exposure, returned as a set of de-noised calibration lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three (\textsection \ref{sec:interp_wsol}) to construct a full wavelength model that returns wavelength as a function of detector position $x$ and echelle order $m$.}
\label{fig:cartoon}
\end{figure*}

@@

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
In \name\, the amplitude, $a_{n,k}$, of each principal component is then interpolated to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  These new, linearly interpolated magnitudes, $a_{n\,k}$ for a science exposure $n'$ at time $t_{n'}$ are then used to construct the calibration state of the spectrograph for that point in time.  Using interpolated magnitudes, $a_{n,k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, a new set of calibration lines, $x(\lambda,m,n')$ can be constructed
 for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
From the de-noising step, \name\ can now construct a set of calibration lines, $x(\lambda,m,n')$ for any exposure $n'$.  To construct a wavelength solution, we invert $x(\lambda,m,n')$ to $\lambda(x,m,n')$ by interpolating known wavelengths of the calibration lines over detector position.  For instance, interpolating the known wavelengths vs. line centers onto every integer x will generate wavelengths for each pixel in an echelle order.

We found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity allows the wavelength solution, $\lambda(x,m,n')$, to be invertible and prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

\section{Data} \label{sec:data}
We tested \name\ using data from \expres, the EXtreme PRecison Spectrograph.  \expres\ is an environmentally-stabilized, fiber-fed optical spectrograph with a median resolving power $R=\lambda/\delta_{\lambda}=\sim137,000$, over the wavelength range of $390-780\, nm$ \citep{jurgenson2016, blackman2020}.  \expres\ has two different wavelength calibration sources, a ThAr lamp and a Menlo Systems laser frequency comb (LFC).  LFCs are unique in that the wavelengths of their emission lines are stable and exactly known to the order of pico-meters  \citep{wilken2012, molaro2013, probst2014}.

Rather than using a simultaneous calibration fiber, two to three LFC exposures are obtained roughly every 30 minutes while the telescope is slewing to the next target.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, so that calibration light travels down the same optical pathway and is projected onto the same pixels as all science observations.

LFC lines cover echelle orders 84-124, which contain approximately 19200 calibration lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include approximately 5300 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.  A list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$, which have been calculated by the EXPRES pipeline \citep{petersburg2020}, are read into \name.

Line positions from the \expres\ pipeline are generated as follows.  A thorium argon (ThAr) wavelength solution is generated from each ThAr exposure using the IDL code \texttt{thid.pro}, developed by Jeff Valenti.  This code identifies ThAr lines by matching lines in an exposure against a line atlas.  Line matching is carried out in an automated and unsupervised way with a Levenburg-Marquardt minimization routine.  Once each line's position is identified and matched to a wavelength from the line atlas, a sixth-order, 2D polynomial is fit over pixel location $x$, echelle order $m$, and scaled wavelength $m\lambda$ (wavelengths are scaled in order to distinguish lines that may appear in more than one order).

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit with a Gaussian.  The mean of this fitted Gaussian to a single peak is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from design of the LFC, and the offset frequency, $f_0$, has been determined by Menlo Systems, the manufacturer of the LFC.

In order to satisfy the assumption that there exists only low-order variation, which is needed for \name, we used exposures from after the photonic crystal fiber was changed in Summer 2019 and the LFC stabilized.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2019 on 29 unique nights.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all tests, we leave out a subset of calibration lines with known wavelengths as a ``validation'' sample, generate wavelengths for these lines using the remaining data, and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but this contribution to the residuals will be the same across all tests.

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even- and odd-indexed lines.  We then construct a wavelength solution using only the odd-indexed lines and use it to predict wavelengths for the even-indexed lines; i.e. a polynomial is fit to just the odd-indexed lines and then evaluated at the detector position of the even-indexed lines (see Equation \ref{eq:poly_wsol}).  We then generate a wavelength solution using only the even-indexed lines and use it to predict wavelengths for the odd-indexed lines.

@@

Resultant wavelengths are compared on a line-by-line basis.  The residuals of a wavelength solution represent the difference between the wavelength solution evaluated at the line position of a calibration line, and the assigned theoretical wavelength (i.e. from Equation \ref{eq:lfc} for LFC lines and from a line atlas for ThAr lines) for each line in every exposure.  The reported RMS of a wavelength solution is therefore the per-line RMS, i.e.
\begin{equation}
RMS/line \: [m\,s^{-1}] = \sqrt{\sum_{n=1}^N\sum_{p=1}^P\frac{[ \frac{(\lambda_{n,p,predicted} - \lambda_{p,theoretical})}{\lambda_{p,theoretical}} \times c ]^2}{N \times P}}
\label{eq:rms}
\end{equation}
where $\lambda_{p,theoretical}$ is the theoretical wavelength for line $p$, $\lambda_{n,p,predicted}$ is the wavelength predicted by the constructed wavelength solution for line $p$ in exposure $n$, and residuals from all $P$ lines from all $N$ exposures are used, for a total of $N \times P$ lines.  The difference in wavelength is converted to units of $\mps$, a more intuitive metric for \eprv\ work.

\begin{figure}[t]
\centering
\includegraphics[width=.6\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The per line RMS as defined in Equation \ref{eq:rms} is given in the top-right corner in each method's corresponding color.  \Name\ returns the smallest spread in residuals.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.55\textwidth]{Figures/lineResids2D_col.png}
%\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals of a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for parametric and non-parametric wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  \textbf{Top}: residuals to a classic, polynomial fit.  \textbf{Bottom}: residuals to an interpolated wavelength solution.  High-order structure, i.e. vertical stripes and patchiness, is apparent in the residuals to a polynomial wavelength solution, which assumes smoothness.}
\label{fig:resid2d}
\end{figure}

@@

@@

This shows how the flexibility of an interpolated model can account for high-order instrument or detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  This same flexibility may similarly allow interpolated wavelength solutions to account for position errors in pixel image blocks for different detectors \citep{fischer2016, milakovic2020}.

>>>>

\documentclass[twocolumn,table,xcolor]{aastex63}
\usepackage[ruled,vlined,norelsize]{algorithm2e}

% hogg dealing with his shit
\addtolength{\topmargin}{-0.30in}
\addtolength{\textheight}{0.70in}
%\renewcommand{\twocolumngrid}{} % hahaha!
\setlength{\parindent}{1.1\baselineskip}
\sloppy\sloppypar\raggedbottom\frenchspacing
\shorttitle{excalibur: a wavelength-calibration model}
\shortauthors{zhao, hogg, bedell, fischer}

@@

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments, which feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all accessible calibration states for an instrument.  \Name\ also takes advantage of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities better than any parametric model, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which includes a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  In a typical test case, radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
%\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity (\eprv) programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Two next-generation spectrographs, \expres\ and \espresso, have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, petersburg2020, brewer2020, mascareno2020}.

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.

The move towards 0.1 $\mps$ RV precision, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to identify incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using a segmented polynomial in the dispersion direction, tuned to capture detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

@@

\section{Method} \label{sec:method}
\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  We exaggerate variations in measured line position, changes in calibration space, and interpolation deviations for clarity.  In step one, dimensionality reduction and denoising (\textsection \ref{sec:denoising}), the complete set of line positions for all exposures is analyzed to return a set of $K$ basis vectors, $g_n$, which represent different ways the spectrograph calibration changes.  These basis vectors span the $K$-dimensional calibration space of the spectrograph, which includes all possible wavelength solutions.  In step two (\textsection \ref{sec:interp_time}), the amplitude of each basis vector, $a_{n,k}$, is interpolated to return the calibration state for a specific science exposure, returned as a set of de-noised calibration lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three (\textsection \ref{sec:interp_wsol}) to construct a full wavelength model that returns wavelength as a function of detector position $x$ and echelle order $m$.}
\label{fig:cartoon}
\end{figure*}

The \name\ method is designed to take many calibration images, each containing a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate this information into a full wavelength model applicable to all exposures taken with the instrument.  It operates on the two core ideas; The wavelength solution should be allowed flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \Name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to a 2D, 9\textsuperscript{th}-order polynomial, where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here, the y-dependence is dropped as, in our framing, this dependence is carried by spectral order $m$.  The line position on the detector is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated from the time of calibration exposures to time $t_n$ of a science exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration, science exposures to re-construct the coefficients for a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  The resultant $x(\lambda,m,n)$, a list of calibration line positions for a given exposure, can be regarded as the calibration state of the spectrograph for that exposure.  When this calibration structure is used to deliver a wavelength solution, $x(\lambda,m,n)$ can be inverted back into $\lambda(x,m,n)$ to recover wavelengths for each detector position $x$ and echelle order $m$.

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present one implementation of \expres\ using principal component analysis (PCA) \citep{wiki_pca}.  A PCA is justifiable in the limit where there is very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

@@

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
In \name, the amplitude, $a_{n,k}$, of each principal component is interpolated to determine the calibration state of the spectrograph.  For example, the amplitude can be interpolated with respect to time to recreate the calibration state of the spectrograph at different times.  The choice of what to interpolate against depends on the dominant contribution to variation in the calibration of the instrument.

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  This is discussed more in Section \ref{sec:choice_avt}.  These new, linearly interpolated magnitudes, $a_{n\,k}$ for a science exposure $n'$ at time $t_{n'}$, are then used to construct the calibration state of the spectrograph for that point in time.  Using interpolated amplitudes, $a_{n,k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, a new set of calibration lines, $x(\lambda,m,n')$ can be constructed
 for any exposure as formulated in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
From the de-noising step, \name\ can now construct a set of calibration lines, $x(\lambda,m,n')$ for any exposure $n'$.  To construct a wavelength solution, we invert $x(\lambda,m,n')$ to $\lambda(x,m,n')$ by interpolating known wavelengths of the calibration lines over detector position.  For instance, interpolating the known wavelengths vs. line centers onto every integer $x$ will generate wavelengths for each pixel in an echelle order.

After experiments, we found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a parameterized function, while the enforced monotonicity allows the wavelength solution, $\lambda(x,m,n')$, to be invertible and prevents spurious deviations that may befall a cubic spline.  Choices in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

\section{Data} \label{sec:data}
We tested \name\ using data from \expres, the EXtreme PRecison Spectrograph.  \expres\ is an environmentally-stabilized, fiber-fed optical spectrograph with a median resolving power $R=\lambda/\delta_{\lambda}=\sim137,000$, over a wavelength range of $390-780\, nm$ \citep{jurgenson2016, blackman2020}.  \expres\ has two different wavelength calibration sources, a thorium argon (ThAr) lamp and a Menlo Systems laser frequency comb (LFC).  LFCs are unique in that the wavelengths of their emission lines are stable and exactly known at pico-meter accuracy \citep{wilken2012, molaro2013, probst2014}.

Rather than using a simultaneous calibration fiber, two to three LFC exposures are obtained roughly every 30 minutes while the telescope is slewing to new targets.  ThAr exposures are taken at the beginning and end of each night.  All calibration data are taken through the science fiber, so that calibration light travels down the same optical pathway and is projected onto the same pixels as the science observations.  Light passes through a pupil slicer and double scrambler before being injected into a rectangular fiber, which is fed through a mechanical agitator to ensure modal mixing\citep{petersburg2018}.

LFC lines cover echelle orders 84-124, which contain approximately 19200 calibration lines.  Though our results are primarily based on work with LFC data, there will be some discussion of applications to arc lamps below.   ThAr lines cover all 86 extracted orders of \expres\ (echelle orders 75-160), which include approximately 5300 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, exposures with more than 60\% of expected lines missing were cut from the analysis.  A list of echelle orders $m$, line wavelengths $\lambda$, and pixel positions $x$, which have been calculated by the EXPRES pipeline \citep{petersburg2020} for every line of every exposure, is read into \name.

Line positions from the \expres\ pipeline are generated as follows.  A ThAr wavelength solution is generated from each ThAr exposure using the IDL code \texttt{thid.pro}, developed by Jeff Valenti.  This code identifies ThAr lines by matching lines in an exposure against a line atlas.  Line matching is carried out in an automated and unsupervised way with a Levenburg-Marquardt minimization routine.  Once each line's position is identified and matched to a wavelength from the line atlas, a sixth-order, 2D polynomial is fit over pixel location $x$, echelle order $m$, and scaled wavelength $m\lambda$ (wavelengths are scaled in order to distinguish lines that may appear in more than one order).

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit with a Gaussian.  The mean of this fitted Gaussian to a single peak is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from the design of the LFC, and the offset frequency, $f_0$, has been determined by Menlo Systems, the manufacturer of the LFC.

In order to satisfy the assumption that there exists only low-order variation, which is needed for \name, we used exposures from after the photonic crystal fiber within the LFC was changed in Summer 2019, and the LFC stabilized.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2019 on 29 unique nights.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, non-hierarchical, parametric method.  To implement training-validation tests, we leave out a subset of calibration lines with known wavelengths as a ``validation'' sample, generate wavelengths for these lines using the remaining data, and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but this contribution to the residuals will be the same across all tests.

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even- and odd-indexed lines.  We then construct a wavelength solution using only the odd-indexed lines and use  that wavelength solution to predict wavelengths for the even-indexed lines; i.e. a polynomial is fit to just the odd-indexed lines and then evaluated at the detector positions of the even-indexed lines (see Equation \ref{eq:poly_wsol}).  We then generate a wavelength solution using only the even-indexed lines and use it to predict wavelengths for the odd-indexed lines.

@@

Resultant wavelengths are compared on a line-by-line basis.  The residuals of a wavelength solution represent the difference between the wavelength solution evaluated at the line position of a calibration line, and the assigned theoretical wavelength (i.e. from Equation \ref{eq:lfc} for LFC lines and from a line atlas for ThAr lines) for each line in every exposure.  The reported RMS of a wavelength solution is therefore the per-line RMS, i.e.
\begin{equation}
RMS/line \: [m\,s^{-1}] = \sqrt{\sum_{n=1}^N\sum_{p=1}^P\frac{[ \frac{(\lambda_{n,p,pred.} - \lambda_{p,theory})}{\lambda_{p,theory}} \times c ]^2}{N \times P}}
\label{eq:rms}
\end{equation}
where $\lambda_{p,theory}$ is the theoretical wavelength for line $p$, $\lambda_{n,p,pred.}$ is the wavelength predicted by the constructed wavelength solution for line $p$ in exposure $n$, and residuals from all $P$ lines from all $N$ exposures are used, for a total of $N \times P$ lines.  The difference in wavelength is converted to units of $\mps$, a more intuitive metric for \eprv\ work.

@@

\begin{figure}[t]
\centering
\includegraphics[width=.45\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The per line RMS as defined in Equation \ref{eq:rms} is given in the top-right corner in each method's corresponding color.  \Name\ returns the smallest spread in residuals.}
\label{fig:testHists}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.455\textwidth]{Figures/lineResids2D_col.png}
%\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals of a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for parametric and non-parametric wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  \textbf{Top}: residuals to a classic, polynomial fit.  \textbf{Bottom}: residuals to an interpolated wavelength solution.  High-order structure, i.e. vertical stripes and patchiness, is apparent in the residuals to a polynomial wavelength solution, which assumes smoothness.}
\label{fig:resid2d}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineDensity.png}
\caption{Residuals when using ThAr lines to predict wavelengths for LFC lines.  \textbf{Right}: residuals for a single LFC exposure plotted with respect to detector position and colored by residual, as in Figure \ref{fig:resid2d}.  The positions of ThAr lines are over-plotted in yellow.  In general, residuals are greater between ThAr lines with greater separation.
\textbf{Left}: Comparison of polynomial and interpolated wavelength solutions using either just ThAr lines or LFC lines for a subset of echelle order 94.  The shape of the residuals from a polynomial fit are similar whether using ThAr lines or LFC lines.  A PCHIP interpolated wavelength model guided by LFC lines returns the smallest residuals.}
\label{fig:waveResids}
\end{figure*}

@@

This shows how the flexibility of an interpolated model can account for high-order instrument or detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  This same flexibility may similarly allow interpolated wavelength solutions to account for position errors in pixel image blocks for different detectors depending on how the interpolation is framed \citep{fischer2016, milakovic2020}.

Though the interpolated wavelength solution returns lower, less-structured residuals than the polynomial wavelength solution when guided by LFC lines, the flexibility of an interpolated wavelength solution can result in much worse residuals when not properly constrained, for example in regions between widely separated calibration lines.  The left plot of Figure \ref{fig:waveResids} shows the residuals when ThAr calibration lines, which are much fewer and less regularly-spaced than LFC lines, are run through \name\ and used to predict wavelengths for the (completely independent) LFC exposures taken during the same range of time.  Over-plotted in yellow are the positions of the ThAr lines.
